---
title: "class09HalloweenCandyProject"
author: "Nashed A16631132"
format: pdf
---

```{r}
candy_file <- "candy-data.csv"
candy = read.csv(candy_file, row.names=1)
head(candy)
```

>Q1. How many different candy types are in this dataset?
85

```{r}
nrow(candy)
```

```{r}
sum(candy$fruity)
```
>Q2. How many fruity candy types are in the data set? 
38

I can convert the 1 and 0 values to be TRUE or FLASE and use that to extract the type of candy I want. For example the chocolate candy....

```{r}
candy[as.logical(candy$chocolate),]
```


>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["3 Musketeers", ]$winpercent
```
>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```
> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```
```{r}
library("skimr")
skim(candy)
```
> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?
Winpercent seems to be on a different scale and has very different values than the other valuables and is not in the 0-1 range. 
>Q7. What do you think a zero and one represent for the candy$chocolate column?
0 is false and 1 true. 

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) + 
  geom_histogram(binwidth=10)
```
>Q9. Is the distribution of winpercent values symmetrical?
No, it is not very symmetrical and is skewed. 

>Q10. Is the center of the distribution above or below 50%?
The center of the distribution is below 50% and is skewed to the right. 


```{r}
choc.inds <- as.logical(candy$chocolate)
choc.win <- candy[choc.inds, "winpercent"]
choc.win
```

```{r}
fruit.inds <- as.logical(candy$fruit)
fruit.win <- candy[fruit.inds, "winpercent"]
fruit.win
```
```{r}
mean(choc.win)
```

```{r}
mean(fruit.win)
```
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?
On average, chocolate is higher ranked than fruit candy and has a higher average. 

```{r}
t.test(choc.win, fruit.win)
```
>Q12. Is this difference statistically significant?
The p-value is very small so we can reject the null hypothesis that there is no difference in the means, and conclude that there is a significant difference. 

```{r}
mycols <- rep("gray", nrow(candy))
#mycols[2:5] <- "red"
mycols[as.logical(candy$chocolate)] <-"chocolate"
mycols[as.logical(candy$fruit)] <-"pink"
mycols
```


>Q13. What are the five least liked candy types in this set?
Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, Jawbusters

```{r}
head(candy[order(candy$winpercent),], n=5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?
Reeses Peanut Butter Cup, Reeses Miniatures, Twix, Kit Kat, Snickers

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.
>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=mycols)
```


```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=mycols) + 
  geom_text_repel(col=mycols, size=3.3, max.overlaps = 5)
```


>Q17. What is the worst ranked chocolate candy?
Nik L Nip
>Q18. What is the best ranked fruity candy?
Reeses Peanut Butter Cup


```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
Reeses miniatures 

>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?
The top 5 most expensive candy typea are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate and the least popular one is Nik L Nip. 

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?
fruity and chocolate

>Q23. Similarly, what two variables are most positively correlated?
winpercent and chocolate

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2], col=mycols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=mycols)
```

```{r}
library(ggrepel)
p + geom_text_repel(size=3.3, col=mycols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```
```{r}
library(plotly)
#ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")

```

>Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?
Fruity, hard, and pluribus because I would expect that fruit candy would come in a bag or box of multiple candies.  